Concrete Semantics for Pushdown Analysis:\\The Essence of Summarization

### 2013-10-04 content.tex
Filled out wcm semantics more. I think markset is unsound in the circular case. Unsure how to fix.

We saw in \autoref{sec:gc} that a semantics can inspect the continuation table to compute a property of the ``whole stack,'' which is outside the model of pushdown systems.
%
Here we show that Clements and Felleisen's WCM machine (a generalization of the CM machine) is directly analyzable in this setting, though we give up some precision in order to account for all calling contexts instead of the single one a function is currently in.
%
The machine can extend the marks of the continuation, reify the current set of marks, and query the set for all the values given to a particular mark key.
%
The mark set of the continuation is then part of the ``context'' for function calls, since functions can inspect the marks.
%
Mark sets can have unbounded length chains of values for each mark key, so we use $\alloc$ to tune the precision of context comparisons.
%
The effect is that we reify the current mark set for each function call to keep as part of the context (one can devise caching strategies for better performance).
%
We use a private store to represent the chains of values, in order to not affect the precision of reifications that happen within the analyzed program.
%
\begin{align*}
  \mexpr \in \Expr &::= \swcm{\mexpr}{\mexpr}{\mexpr} \alt \sccms \alt \scmsl{\mexpr}{\mexpr} \alt \ldots \\
  \mctx \in \Context &::= ((\mexpr, \menv), \mstore, \mmarkset) \\
  \mkframe \in \Frame &::= \kwcmk{\mexpr, \mexpr, \menv} \alt
                           \kwcmv{\mval, \mexpr, \menv} \alt
                           \kcmsls{\mexpr,\menv} \alt
                           \kcmslk{\mval} \alt \ldots \\ 
  \mkont \in \Kont &::= \kmt^\mmarks \alt \krt{\mctx} \alt \kcons{\mkframe^\mwmarks}{\mkont} \\
  \mmarks \in \Marks &= \Value \to \wp(\Value) \\
  \mwmarks \in \QualifiedMarks &::= \scopy{\mmarks} \alt \sext{\mmarks} \\
  \mmarkset \in \Markset &= \Marks \times \Store \\
  \mval \in \Value &::= \texttt{nil} \alt \cons{\maddr}{\maddr} \alt \smarkset{\mmarks} \alt \ldots
\end{align*}
%% TODO: distinguish between copied and extended marks so that domarkset only builds
%% a list of mark extensions
\begin{figure}
  \centering
  $\mstate \stepto \mstate' \text{ where } \maddr = \alloc(\mstate)$ \\
  \begin{tabular}{r|l}
    \hline
% wcm eval key
    $\tpl{(\swcm{\mexpr_k}{\mexpr_v}{\mexpr}, \menv), \mstore, \mkont, \mktab, \mmemo}$
    &
    $\tpl{(\mexpr_k, \menv), \mstore, \kcons{\kwcmk{\mexpr_v,\mexpr,\menv}}{\mkont}, \mktab, \mmemo}$
    \\
% wcm eval value
    $\tpl{\mval_k, \mstore, \kcons{\kwcmk{\mexpr_v,\mexpr,\menv}}{\mkont}, \mktab, \mmemo}$
    &
    $\tpl{(\mexpr_v, \menv), \mstore, \kcons{\kwcmv{\mval_k,\mexpr,\menv}}{\mkont}, \mktab, \mmemo}$
    \\    
% wcm all done.
    $\tpl{\mval_v, \mstore, \kcons{\kwcmv{\mval_k, \mexpr, \menv}}{\mkont}, \mktab, \mmemo}$
    &
    $\tpl{\mexpr,\menv,\mstore,\mkont',\mktab', \mmemo}$ \\
    where & $(\mkont',\mktab') = \domark(\mkont,\mktab,\mval_k,\mval_v)$
    \\
% function call
    $\tpl{\mval,\mstore,\kcons{\kfn{\vclo{\slam{\mvar}{\mexpr}}{\menv}}}{\mkont}}$
    &
    $\tpl{(\mexpr, \menv'), \mstore', \krt{\mctx},\mktab',\mmemo}$
    \\ where & $\menv' = \extm{\menv}{\mvar}{\maddr}$ \\
    & $\mstore' = \joinone{\mstore}{\maddr}{\mval}$ \\
    & $\mctx = ((\mexpr,\menv'),\mstore', \domarkset(\mstate))$ \\
    & $\mktab' = \joinone{\mktab}{\mctx}{\mkont}$
    \\
% Current continuation marks
    $\tpl{(\sccms,\menv),\mstore,\mkont,\mktab,\mmemo}$
    &
    $\tpl{\smarkset{\mmarks},\mstore\sqcup\mstore',\mkont,\mktab,\mmemo}$
    \\ where & $(\mmarks,\mstore') = \domarkset(\mstate)$
    \\
% cms->list mark set
    $\tpl{(\scmsl{\mexpr_\mmarks}{\mexpr_k}, \menv), \mstore, \mkont, \mktab, \mmemo}$
    &
    $\tpl{(\mexpr_\mmarks, \menv), \mstore, \kcons{\kcmsls{\mexpr_k,\menv}}{\mkont}, \mktab, \mmemo}$
    \\
% cms->list key
    $\tpl{\mval, \mstore, \kcons{\kcmsls{\mexpr_k,\menv}}{\mkont}, \mktab, \mmemo}$
    &
    $\tpl{(\mexpr_k, \menv), \mstore, \kcons{\kcmslk{\mval}}{\mkont}, \mktab, \mmemo}$
    \\
% cms->list done
    $\tpl{\mval_k, \mstore, \kcons{\kcmslk{\mmarks}}{\mkont}, \mktab, \mmemo}$
    &
    $\tpl{\mval, \mstore, \mktab, \mmemo}$ if $\mval \in \mmarks(\mval_k)$
  \end{tabular}  
  \caption{The Tabular WCM machine (list rules not shown)}
  \label{fig:wcm}
\end{figure}

The two metafunctions, $\domark$ and $\domarkset$ are what do the heavy-lifting of adding marks to a continuation and reifying the mark set into a map of keys to lists.
%
Adding a mark to a continuation has an implicit equality check; if $\mval_k = \mval_k'$ concretely, then {\tt wcm} should do a strong update on $\mmarks$, but otherwise it should join since they could be two different values.
%
Concrete equality analysis is outside the scope of this paper, but doable (\eg, with abstract counting~\citep{dvanhorn:Might:2006:GammaCFA}), so we elide the details and call this strong-update-or-join operation $\sqcup_{!{}}$.
%
Marking a return continuation means we need to crawl through the table, which again can be circular, so we compute the least fixed point of ${\mathcal M}$ to get an extended table holding the marked continuations.
%
\newcommand{\rewritectx}{\mathit{rw}}
\newcommand{\domarkstep}{\mathit{mark}_0}
\begin{align*}
  \domark(\mkont, \mktab, \mval_k, \mval_v) &= (\mkont', \mktab) \text{ if } (\mkont',\bot) = \domarkstep(\mkont) \\
  \domark(\mkont, \mktab, \mval_k, \mval_v) &= (\mkont', \mathbf{snd}(\lfp({\mathcal M}(\mctx)))) \text{ if } (\mkont',\mctx) = \domarkstep(\mkont) \\
  \text{where } {\mathcal M}(\mctx_0)(C, \mktab) &= (C \cup C' \cup \set{\mctx_0}, \mktab \sqcup \mktab') \\
    I &= \setbuild{(\rewritectx(\mctx), \setbuild{\domarkstep(\mkont')}{\mkont' \in \mktab(\mctx)})}{\mctx \in C} \\
    C' &= \setbuild{\mctx}{(\_, S) \in I, (\_, \mctx) \in S} \\
    \mktab' &= \bigsqcup\setbuild{[\mctx \mapsto \setbuild{\mkont}{(\mctx, S) \in I, (\mkont, \_) \in S}]}{(\mctx,\_) \in I}
\\[2pt]
  \rewritectx((\mexpr,\menv),\mstore,(\mmarks, \mstore')) &= ((\mexpr,\menv),\mstore,(\mmarks\sqcup_{!{}}[\mval_k \mapsto \mval_v], \mstore'))
\\[2pt]
  \domarkstep(\krt{\mctx}) &= (\krt{\rewritectx(\mctx)}, \mctx) \\
  \domarkstep(\kcons{\mkframe^\mmarks}{\mkont}) &= (\kcons{\mkframe^{\mmarks \sqcup_{!{}} [\mval_k \mapsto \mval_v]}}{\mkont}, \bot) \\
  \domarkstep(\kmt^\mmarks) &= (\kmt^{\mmarks \sqcup_{!{}} [\mval_k \mapsto \mval_v]}, \bot)
\end{align*}

Second, the $\domarkset$ metafunction builds a map of keys to lists, and builds a private store for the list constructions.
%
For this, we extend $\alloc$'s specification to allow a second argument, the private store, and produce a pair of addresses.
%
%% FIXME: need to crawl through qualified continuations for all extensions to build a list for each found key.
%% Has another implicit equality check
\begin{align*}
  \domarkset(\mstate\equiv\tpl{\mpoint,\mstore,\mkont,\mktab,\mmemo}) &=
    (\maddr_a, \maddr_d) = \alloc(\mstate,\mstore')
\end{align*}
}

>>>>

We saw in \autoref{sec:gc} that a semantics can inspect the continuation table to compute a property of the ``whole stack,'' which is outside the model of pushdown systems.
%
Here we show that Clements and Felleisen's WCM machine (a generalization of the CM machine) is directly analyzable in this setting, though we give up some precision in order to account for all calling contexts instead of the single one a function is currently in.
%
The machine can extend the marks of the continuation, reify the current set of marks, and query the set for all the values given to a particular mark key.
%
The mark set of the continuation is then part of the ``context'' for function calls, since functions can inspect the marks.
%
Mark sets can have unbounded length chains of values for each mark key, so we use $\alloc$ to tune the precision of context comparisons.
%
The effect is that we reify the current mark set for each function call to keep as part of the context (one can devise caching strategies for better performance).
%
We use a private store to represent the chains of values, in order to not affect the precision of reifications that happen within the analyzed program.
%
\begin{align*}
  \mexpr \in \Expr &::= \swcm{\mexpr}{\mexpr}{\mexpr} \alt \sccms \alt \scmsl{\mexpr}{\mexpr} \alt \ldots \\
  \mctx \in \Context &::= ((\mexpr, \menv), \mstore, \mmarkset) \\
  \mkframe \in \Frame &::= \kwcmk{\mexpr, \mexpr, \menv} \alt
                           \kwcmv{\mval, \mexpr, \menv} \alt
                           \kcmsls{\mexpr,\menv} \alt
                           \kcmslk{\mval} \alt \kwcm{\mmarks} \alt \ldots \\
  \mkont \in \Kont &::= \kmt^\bot \alt \krt{\mctx}^\mmarks \alt \kcons{\mkframe^\mmarks}{\mkont} \\
  \mmarks \in \Marks &= \Value \to \wp(\Value) \\
  \mmarkset \in \Markset &= \Marks \times \Store \\
  \mval \in \Value &::= \texttt{nil} \alt \cons{\maddr}{\maddr} \alt \smarkset{\mmarks} \alt \ldots
\end{align*}

%% TODO: distinguish between copied and extended marks so that domarkset only builds
%% a list of mark extensions
\begin{figure}
  \centering
  $\mstate \stepto \mstate' \text{ where } \maddr = \alloc(\mstate)$ \\
  \begin{tabular}{r|l}
    \hline
% wcm eval key
    $\tpl{(\swcm{\mexpr_k}{\mexpr_v}{\mexpr}, \menv), \mstore, \mkont^\mmarks, \mktab, \mmemo}$
    &
    $\tpl{(\mexpr_k, \menv), \mstore, \kcons{\kwcmk{\mexpr_v,\mexpr,\menv}^\mmarks}{\mkont}, \mktab, \mmemo}$
    \\
% wcm eval value
    $\tpl{\mval_k, \mstore, \kcons{\kwcmk{\mexpr_v,\mexpr,\menv}}{\mkont^\mmarks}, \mktab, \mmemo}$
    &
    $\tpl{(\mexpr_v, \menv), \mstore, \kcons{\kwcmv{\mval_k,\mexpr,\menv}^\mmarks}{\mkont}, \mktab, \mmemo}$
    \\    
% wcm all done.
    $\tpl{\mval_v, \mstore, \kcons{\kwcmv{\mval_k, \mexpr, \menv}}{\mkont}, \mktab, \mmemo}$
    &
    $\tpl{\mexpr,\menv,\mstore,\mkont',\mktab', \mmemo}$ \\
    where & $(\mkont',\mktab') = \domark(\mkont,\mktab,\mval_k,\mval_v)$
    \\
% Pop marks
   $\tpl{\mval,\mstore,\kcons{\kwcm{\mmarks}^{\mmarks'}}{\mkont}, \mktab, \mmemo}$
   &
   $\tpl{\mval,\mstore,\mkont,\mktab,\mmemo}$
   \\
% function call
    $\tpl{\mval,\mstore,\kcons{\kfn{\vclo{\slam{\mvar}{\mexpr}}{\menv}}^\mmarks}{\mkont}}$
    &
    $\tpl{(\mexpr, \menv'), \mstore', \krt{\mctx}^\mmarks,\mktab',\mmemo}$
    \\ where & $\menv' = \extm{\menv}{\mvar}{\maddr}$ \\
    & $\mstore' = \joinone{\mstore}{\maddr}{\mval}$ \\
    & $\mctx = ((\mexpr,\menv'),\mstore', \domarkset(\mstate))$ \\
    & $\mktab' = \joinone{\mktab}{\mctx}{\mkont}$
    \\
% Current continuation marks
    $\tpl{(\sccms,\menv),\mstore,\mkont,\mktab,\mmemo}$
    &
    $\tpl{\smarkset{\mmarks},\mstore\sqcup\mstore',\mkont,\mktab,\mmemo}$
    \\ where & $(\mmarks,\mstore') = \domarkset(\mstate)$
    \\
% cms->list mark set
    $\tpl{(\scmsl{\mexpr_\mmarks}{\mexpr_k}, \menv), \mstore, \mkont^\mmarks, \mktab, \mmemo}$
    &
    $\tpl{(\mexpr_\mmarks, \menv), \mstore, \kcons{\kcmsls{\mexpr_k,\menv}^\mmarks}{\mkont}, \mktab, \mmemo}$
    \\
% cms->list key
    $\tpl{\mval, \mstore, \kcons{\kcmsls{\mexpr_k,\menv}}{\mkont^\mmarks}, \mktab, \mmemo}$
    &
    $\tpl{(\mexpr_k, \menv), \mstore, \kcons{\kcmslk{\mval}^\mmarks}{\mkont}, \mktab, \mmemo}$
    \\
% cms->list done
    $\tpl{\mval_k, \mstore, \kcons{\kcmslk{\mmarks}}{\mkont}, \mktab, \mmemo}$
    &
    $\tpl{\mval, \mstore, \mktab, \mmemo}$ if $\mval \in \mmarks(\mval_k)$
  \end{tabular}  
  \caption{The Tabular WCM machine (list rules not shown)}
  \label{fig:wcm}
\end{figure}

The two metafunctions, $\domark$ and $\domarkset$ are what do the heavy-lifting of adding marks to a continuation and reifying the mark set into a map of keys to lists.
%
Adding a mark to a continuation has an implicit equality check; if $\mval_k = \mval_k'$ concretely, then {\tt wcm} should do a strong update on $\mmarks$, but otherwise it should join since they could be two different values.
%
Concrete equality analysis is outside the scope of this paper, but doable (\eg, with abstract counting~\citep{dvanhorn:Might:2006:GammaCFA}), so we elide the details and call this strong-update-or-join operation $\sqcup_{!{}}$.
%
Marking a return continuation means we need to crawl through the table, which again can be circular, so we compute the least fixed point of ${\mathcal M}$ to get an extended table holding the marked continuations.
%
\newcommand{\rewritectx}{\mathit{rw}}
\newcommand{\domarkstep}{\mathit{mark}_0}
\begin{align*}
  \domark(\mkont, \mktab, \mval_k, \mval_v) &= (\mkont', \mktab) \text{ if } (\mkont',\bot) = \domarkstep(\mkont) \\
  \domark(\mkont, \mktab, \mval_k, \mval_v) &= (\mkont', \mathbf{snd}(\lfp({\mathcal M}(\mctx)))) \text{ if } (\mkont',\mctx) = \domarkstep(\mkont) \\
  \text{where } {\mathcal M}(\mctx_0)(C, \mktab) &= (C \cup C' \cup \set{\mctx_0}, \mktab \sqcup \mktab') \\
    I &= \setbuild{(\rewritectx(\mctx), \setbuild{\domarkstep(\mkont')}{\mkont' \in \mktab(\mctx)})}{\mctx \in C} \\
    C' &= \setbuild{\mctx}{(\_, S) \in I, (\_, \mctx) \in S} \\
    \mktab' &= \bigsqcup\setbuild{[\mctx \mapsto \setbuild{\mkont}{(\mctx, S) \in I, (\mkont, \_) \in S}]}{(\mctx,\_) \in I}
\\[2pt]
  \rewritectx((\mexpr,\menv),\mstore,(\mmarks, \mstore')) &= ((\mexpr,\menv),\mstore,(\mmarks\sqcup_{!{}}[\mval_k \mapsto \mval_v], \mstore'))
\\[2pt]
  \domarkstep(\krt{\mctx}^\mmarks) &= (\krt{\rewritectx(\mctx)}^{\mmarks \sqcup_{!{}} [\mval_k \mapsto \mval_v]}, \mctx) \\
  \domarkstep(\kcons{\kwcm{\mmarks}^{\mmarks'}}{\mkont}) &= (\kcons{\kwcm{\mmarks \sqcup_{!{}} [\mval_k \mapsto \mval_v]}^{\mmarks' \sqcup_{!{}} [\mval_k \mapsto \mval_v]}}{\mkont}, \bot) \\
  \domarkstep(\mkont^\mmarks) &= (\kcons{\kwcm{[\mval_k \mapsto \mval_v]}^{\mmarks \sqcup_{!{}} [\mval_k \mapsto \mval_v]}}{\mkont}, \bot)
\end{align*}

Second, the $\domarkset$ metafunction builds a map of keys to lists, and builds a private store for the list constructions.
%
For this, we extend $\alloc$'s specification to allow more arguments --- the private store, and a set of keys that need cons cells allocated for them --- and produce a map from keys to pairs of addresses.
%
%% FIXME: need to crawl through qualified continuations for all extensions to build a list for each found key.
%% Has another implicit equality check
\newcommand{\markaux}{\mathit{marks}}
\begin{align*}
  \domarkset(\mstate\equiv\tpl{\mpoint,\mstore,\mkont,\mktab,\mmemo}) &= \markaux(\mkont,\emptyset) \\
  \markaux(\kmt^\bot, G) &= (\bot, \bot) \\
  \markaux(\krt{\mctx}^\mmarks, G) &= (\bot, \bot) \text{ if } \mctx \in G \\ %% FIXME: unsound
  \markaux(\krt{\mctx}^\mmarks, G) &= \bigsqcup\limits_{\mkont\in \mktab(\mctx)}{\markaux(\mkont,G \cup \set{\mctx})} \text{ otherwise} \\
  \markaux(\kcons{\kwcm{\mmarks}^{\mmarks'}}{\mkont}, G) &= (\mmarks''[\mval_k \mapsto \set{\cons{a}{d}} : \mval_k \in \dom(\mmarks), (a, d) = A(\mval_k)], \\
    &\phantom{= (}
      \mstore\sqcup[a \mapsto \mmarks(\mval_k) : \mval_k \in \dom(\mmarks), (a,\_) = A(\mval_k)] \\
    &\phantom{= (\mstore}\sqcup[d \mapsto \mmarks''(\mval_k) : \mval_k \in \dom(\mmarks), (\_,d) = A(\mval_k)]
   \\ \text{where } (\mmarks'', \mstore) &= \markaux(\mkont, G)
   \\ A &= \alloc(\mstate, \mstore, \dom(\mmarks))
\end{align*}
}
